---
title: "HW1"
author: "Michelle Hernandez"
format:
  html:
    embed-resources: true
editor: visual
---

```{r}
library(data.table)
library(leaflet)
library(ggplot2)
library(dplyr)
library(DescTools)
```

Read in data

```{r}
#| echo: false
read_csv_individual = read.csv('/Users/michellehernandez/Desktop/DataScience/HW2/HW2/chs_individual.csv')
individual_dat = data.table(read_csv_individual)
#dimensions for individual_dat 1200 x 23
read_csv_region = read.csv('/Users/michellehernandez/Desktop/DataScience/HW2/HW2/chs_regional.csv')
region_dat = data.table(read_csv_region)

#dimensions for read_csv_individual 1200 x 23
```

```{r}
#| echo: false
length(individual_dat$sid)
# 1200 , there are 1200 subjects
str(region_dat)
#Classes ‘data.table’ and 'data.frame':	12 obs. of  27 variables

str(individual_dat)
#Classes ‘data.table’ and 'data.frame':	1200 obs. of  23 variables:
```

merge & explore data. We can see after merging that there are 1200 rows and 49 variables as expected.

```{r}

dat <- merge(individual_dat, region_dat, by= 'townname')
str(dat)
# Classes ‘data.table’ and 'data.frame':	1200 obs. of  49 variables

table(dat$townname)
# There are 100 observations for each of the 12 towns
table(dat$asthma) #Most people in the study population don't have asthma (998 without asthma vs 171 with asthma)
table(dat$male) # We have a relatively even spread of males and females (610 females, 590 males)

hist(dat$fev) # FEV is normally distributed with a mean FEV of 2031.3
summary(dat$fev)
hist(dat$pm25_mass) # there is only 1 pm25_mass values for each town
unique(dat$pm25_mass)
# There are 12 unique values of pM25_mass, 12 unique towns for 1 value for each town 
summary(dat$pm25_mass)
```

Fromt the output above we can confirmt that the data merged correctly. The new data set, dat, has 1200 obs. of 49 variables which is what we expected. There are 100 observations for each of the 12 towns. Most people in the study population don't have asthma (998 without asthma vs 171 with asthma) We have a relatively even spread of males and females (610 females, 590 males). FEV is normally distributed with a mean FEV of 2031.3. There is only 1 pm25_mass values for each town. There are 12 unique values of pM25_mass, 12 unique towns for 1 value for each town.

### Imputation

We want to have 4 imputation groups. 1 for hispanic males, 1 for hispanic females, 1 for non hispanic males, and 1 for non Hispanic females. For continous variables we will take the mean for each group and impute any NAs within their respective groups. We will do the same for categorical variables expect we will use mode.

dat\[, temp_imp := fcoalesce(temp, mean(temp, na.rm = TRUE)), by = .(STATE, year, month, day)\]

```{r}
# These are the groups we want to impute off of
dat[, Mode(mother_asthma, na.rm = TRUE), by = .(hispanic, male)]

dat[,mean(fev, na.rm = TRUE), by = .(hispanic, male)]
```

```{r}
#example for continuous
dat[, fev := fcoalesce(fev, mean(fev, na.rm = TRUE)),
    by = .(hispanic, male)]
# categorical
dat[, mother_asthma := fcoalesce(mother_asthma, Mode(mother_asthma, na.rm = TRUE)),by = .(hispanic, male)]
#dat
```

```{r}
# IMPUTE
#| echo: false
#categorical 
dat[, asthma := fcoalesce(asthma, Mode(asthma, na.rm = TRUE)), by = .(hispanic, male)]
dat[, active_asthma := fcoalesce(active_asthma, Mode(active_asthma, na.rm = TRUE)), by = .(hispanic, male)]
dat[, father_asthma := fcoalesce(father_asthma, Mode(father_asthma, na.rm = TRUE)), by = .(hispanic, male)]
dat[, mother_asthma := fcoalesce(mother_asthma, Mode(mother_asthma, na.rm = TRUE)), by = .(hispanic, male)]
dat[, wheeze := fcoalesce(wheeze, Mode(wheeze, na.rm = TRUE)), by = .(hispanic, male)]
dat[, hayfever := fcoalesce(hayfever, Mode(hayfever, na.rm = TRUE)), by = .(hispanic, male)]
dat[, allergy := fcoalesce(allergy, Mode(allergy, na.rm = TRUE)), by = .(hispanic, male)]
dat[, educ_parent := fcoalesce(educ_parent, Mode(educ_parent, na.rm = TRUE)), by = .(hispanic, male)]
dat[, smoke := fcoalesce(smoke, Mode(smoke, na.rm = TRUE)), by = .(hispanic, male)]
dat[, pets := fcoalesce(pets, Mode(pets, na.rm = TRUE)), by = .(hispanic, male)]
dat[, gasstove := fcoalesce(gasstove, Mode(gasstove, na.rm = TRUE)), by = .(hispanic, male)]


#continuous
dat[, agepft := fcoalesce(agepft, mean(agepft, na.rm = TRUE)), by = .(hispanic, male)]
#dat[, height := fcoalesce(height, mean(height, na.rm = TRUE)), by = .(hispanic, male)]
dat[, height := fcoalesce(as.double(height), mean(as.double(height), na.rm = TRUE)), by = .(hispanic, male)]

dat[, weight := fcoalesce(as.double(weight), mean(as.double(weight), na.rm = TRUE)), by = .(hispanic, male)]
dat[, bmi := fcoalesce(bmi, mean(bmi, na.rm = TRUE)), by = .(hispanic, male)]
dat[, fev := fcoalesce(fev, mean(fev, na.rm = TRUE)), by = .(hispanic, male)]
dat[, fvc := fcoalesce(fvc, mean(fvc, na.rm = TRUE)), by = .(hispanic, male)]
dat[, mmef := fcoalesce(mmef, mean(mmef, na.rm = TRUE)), by = .(hispanic, male)]

```

### create bmi categories

```{r}

dat <- dat %>%
  mutate(
    obesity_level = case_when(
           bmi < 14 ~ 'underweight',
           bmi >= 14 & bmi < 22 ~ 'normal',
           bmi >=22 & bmi < 24 ~ 'overweight',
           bmi >= 24 ~ 'obese',
           TRUE ~ 'NA'
         )) %>%
  collect()

```

To make sure the variable is rightly coded, create a summary table that contains the minimum BMI, maximum BMI, and the total number of observations per category:

```{r}
summary_table <- dat %>%
  group_by(obesity_level) %>%
  summarize(
    min_bmi = min(bmi),
    max_bmi = max(bmi),
    total_observations = n()
  )

summary_table
```

### Create smoke/gas exposure variable and table

smoke_gas_exposure

smoke 1: exposed to second hand smoke in home , 0: not exposed

gas stove 1: presence of gas stove in home 0: not in home

```{r}
dat <- dat %>%
  mutate(
    smoke_gas_exposure = case_when(
           smoke ==1 & gasstove==1 ~ 'exp_both',
           smoke ==1 & gasstove==0 ~ 'exp_smoke',
           smoke ==0 & gasstove==1 ~ 'exp_gas',
           smoke ==0 & gasstove==0 ~ 'exp_none',
           TRUE ~ 'NA'
         )) %>%
  collect()



```

```{r}
summary_exp <- dat %>%
  group_by(smoke_gas_exposure) %>%
  summarize(
    total_observations = n()
  )

summary_exp
```

Create four summary tables showing the average (or proportion, if binary) and sd of "Forced expiratory volume in 1 second (ml)" (an asthma indicator) by town, sex, obesity level, and "smoke_gas_exposure."

```{r}
#Table of average FEV by town
sumtab1 <- dat %>%
  group_by(townname) %>%
  summarize(
    min_fev = min(fev),
    max_fev = max(fev),
    mean_fev_t = mean(fev),
    sd_fev_t = sd(fev),
    total_observations = n()
  )
sumtab1

#Table of gender proportions
sumtab2 <- dat %>%
  summarize(
    total_observations = n(),
    prop_male = sum(male == 1) / n(),
    prop_female = sum(male == 0) / n()
  )
sumtab2
#Table of average FEV by gender
sumtabmale <- dat %>%
  group_by(male) %>%
  summarize(
    min_fev = min(fev),
    max_fev = max(fev),
    mean_fev_t = mean(fev),
    sd_fev_t = sd(fev),
    total_observations = n()
  )
sumtabmale

#Table of average FEV by obesity level
sumtab3 <- dat %>%
  group_by(obesity_level) %>%
  summarize(
    min_fev = min(fev),
    max_fev = max(fev),
    mean_fev_t = mean(fev),
    sd_fev_t = sd(fev),
    total_observations = n()
  )
sumtab3

#Table of average FEV by Smoke/Gas Exposure
sumtab4 <- dat %>%
  group_by(smoke_gas_exposure) %>%
  summarize(
    min_fev = min(fev),
    max_fev = max(fev),
    mean_fev_t = mean(fev),
    sd_fev_t = sd(fev),
    total_observations = n()
  )
sumtab4
```

```{r}
#explaratory 

table(dat$asthma) #Most people in the study population don't have asthma (1029 without asthma vs 171 with asthma)
table(dat$male) # We have a relatively even spread of males and females (610 females, 590 males)


summary(dat$fev)
summary(dat$pm25_mass)
summary(dat$bmi)
summary(dat$smoke_gas_exposure)



hist(dat$fev, main = "Histogram of FEV", xlab = "FEV Values", ylab = "Frequency") # FEV is normally distributed with a mean FEV of 2031.3


ggplot(data = dat, aes(x = townname, y = pm25_mass)) +
  geom_bar(stat = "identity", fill = "darkolivegreen1") +
  labs(x = "Town", y = "PM2.5 Mass") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  +
  ggtitle("Mean PM2.5 Mass by Town")

table(dat$smoke_gas_exposure)
data_counts <- data.frame(
  exposure = c("exp_both", "exp_gas", "exp_none", "exp_smoke"),
  count = c(154, 791, 219, 36)
)
ggplot(data = data_counts, aes(x = exposure, y = count)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(x = "Exposure Type", y = "Count") +
  theme_minimal() +
  ggtitle("Counts of Exposure Types")

hist(dat$bmi, main = "Histogram of BMI", xlab = "BMI Values", ylab = "Frequency")
table(dat$obesity_level)
obesity_counts <- data.frame(
  exposure = c("normal", "obese", "overweight", "underweight"),
  count = c(975, 103, 87, 35)
)
ggplot(data = obesity_counts, aes(x = exposure, y = count)) +
  geom_bar(stat = "identity", fill = "magenta") +
  labs(x = "Obesity Level", y = "Count") +
  theme_minimal() +
  ggtitle("Count of Obesity Levels ")

unique(dat$pm25_mass)
# There are 12 unique values of pM25_mass, 12 unique towns for 1 value for each town 


```

From the graphs above we can see the FEV is normally distributed with a mean FEV of 2031.3. We can also view the distribution of mean pm25_mass by town. Mira Loma has the highest mean pm25 level and Lompoc has the lowest. A large majority of people in the study were exposed to gas, with very few people being exposed to smoke only, and a small portion exposed to both gas and smoke. BMI is also mostly normally distributed with a left skew. There appears to be a few outlines with a higher BMI. When we look at obesity levels we can see most people in the study fall into the normal weight category. With \~100 people falling into the obese category, \~75 in the overweight, and \~ 50 in the underweight category.

### Correlations

```{r}

cont_vars <- c("fev", "bmi", "pm25_mass")  
subset_dat <- dat[, ..cont_vars]
correlation_matrix <- cor(subset_dat)

correlation_matrix


model_obese <- lm(fev ~ obesity_level, data = dat)
summary(model_obese)
model_exp <- lm(fev ~ smoke_gas_exposure, data = dat)
summary(model_exp)
model_pm25 <- lm(fev ~ pm25_mass, data = dat)
summary(model_pm25)
```

From the tests above we can see that FEV and PM25_MASS are significantly associated (p=0.011). BMI and FEV have a positive correlation coefficient of 0.357. FEV and obesity are significantly associated with a p\<0.0001. FEV and smoke/gas exposure were not found to be significantly associated p=0.52.

## **Visualization**

### 1. Facet plot showing scatterplots with regression lines of BMI vs FEV by "townname".

```{r}
ggplot(dat, aes(x=bmi, y=fev)) +
  geom_point() +
  geom_smooth(method="lm") +
  facet_wrap(~townname, nrow=3) +
  labs(x = "BMI", y = "FEV", title = "Relationship Between BMI and FEV by Town")

```

According to the plots it looks like most towns have a slightly positive correlation between BMI and FEV. Riverside, Alpine, and Lompoc have weak positive correlations. This makes sense as Lompoc had the lowest mean PM25_mass.

### 2. Stacked histograms of FEV by BMI category and FEV by smoke/gas exposure. Use different color schemes than the ggplot default.

```{r}
ggplot(dat, aes(x=fev, fill=obesity_level)) +
  geom_histogram() +
  scale_fill_brewer(palette = 'Spectral')+
  labs(x = "FEV Value", y = "Frequency", title = "FEV Distribution by Obesity Level")
```

This graph displays the range of FEV values. The color represents the obesity level categories. We can see that those in underweight category have lower FEV values compared to the other obesity categories. The normal obesity group is centered around the mean FEV, while overweight and obese have higher FEV values.

```{r}
ggplot(dat, aes(x=fev, fill=smoke_gas_exposure)) +
  geom_histogram() +
  scale_fill_brewer(palette = 'PRGn')+
  labs(x = "FEV Value", y = "Frequency", title = "FEV Distribution by Smoke and Gas Exposure")
```

In this graph we are visualizing the distribution of FEV values among different exposure categories. It appears that the FEV distribution is about the same across all categories. This is a similar conclusion that we made from the statistical test above.

### 3. Barchart of BMI by smoke/gas exposure.

```{r}

#boxplot 
ggplot(dat, aes(x = smoke_gas_exposure, y = bmi, fill = smoke_gas_exposure)) +
  geom_boxplot() +
  labs(
    title = "BMI by Smoke/Gas Exposure",
    x = "Smoke/Gas Exposure",
    y = "BMI"
  ) +
  theme_minimal()


# bar chart
mean_bmi <- dat %>%
  group_by(smoke_gas_exposure) %>%
  summarize(mean_bmi = mean(bmi))

ggplot(mean_bmi, aes(x = smoke_gas_exposure, y = mean_bmi, fill = smoke_gas_exposure)) +
  geom_col(position = "dodge") +
  labs(
    title = "Mean BMI by Smoke/Gas Exposure",
    x = "Smoke/Gas Exposure",
    y = "Mean BMI"
  ) +
  theme_minimal()
```

The boxplot shows that there is not a big difference in BMI between the different exposure groups. However exposure to gas and gas+smoke had some higher BMI outliers. The bar graph displays the mean BMI by exposure group, we also see here that there is not a big difference in mean BMI between the exposure groups.

```{r}
mean_bmi <- dat %>%
  group_by(smoke_gas_exposure) %>%
  summarize(mean_bmi = mean(bmi, na.rm = TRUE))
```

### 4. Statistical summary graphs of FEV by BMI and FEV by smoke/gas exposure category.

```{r}
obesity_order <- c("underweight", "normal", "overweight", "obese")
ggplot(dat, aes(x = factor(obesity_level, levels = obesity_order), y = fev)) +
  geom_boxplot() +
  labs(
    title = "FEV by Obesity Levels",
    x = "Obesity Level",
    y = "FEV"
  ) +
  theme_minimal()
```

This boxplot shows the distribution of FEV by obesity level. We can see there is a steady increase of FEV as the obesity level increases. We can see there is a large difference between the median underweight FEV value (around 1700) compared to the obese category (around 2250).

```{r}
ggplot(dat, aes(x = cut(bmi, breaks = seq(0, 50, by = 3)), y = fev)) +
  geom_boxplot() +
  labs(
    title = "FEV by BMI binned",
    x = "BMI Categories",
    y = "FEV"
  ) +
  theme_minimal()
```

This boxplot is similar to the one above, but it has smaller bins for BMI, enabling a more detailed examination of the distribution breakdown. It reveals a consistent upward trend in FEV as BMI increments. With the highest FEV values at the 27-30 BMI range and then the FEV plateaus

```{r}
ggplot(dat, aes(x = bmi, y = fev)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(
    title = "FEV by BMI",
    x = "BMI",
    y = "FEV"
  ) +
  theme_minimal()
```

Here we can see FEV and BMI are positively associated, the FEV increases as BMI increases.

```{r}
exposure_order <- c("exp_none", "exp_gas", "exp_smoke", "exp_both")
ggplot(dat, aes(x = factor(smoke_gas_exposure, levels = exposure_order), y = fev)) +
  geom_boxplot() +
  labs(
    title = "FEV by Smoke/Gas Exposure",
    x = "Smoke/Gas Exposure Category",
    y = "FEV"
  ) +
  theme_minimal()
```

FEV levels are relatively similar across smoke/gas exposure groups. This agrees with our analysis above.

### 5. A leaflet map showing the concentrations of PM2.5 mass in each of the CHS communities.

```{r}

paletteNum <- colorNumeric(c('darkgreen','goldenrod','brown'), domain = dat$pm25_mass)


leaflet(dat) %>% 
  addProviderTiles('CartoDB.Positron') %>% 
  addCircles(lat = ~lat, lng = ~lon, opacity = 1, fillOpacity = 0.8, radius = 400, color = ~paletteNum(dat$pm25_mass)) %>%
  addLegend(
    pal = paletteNum,
    values = ~dat$pm25_mass,
    title = "PM2.5 Mass",
    opacity = 1,
    position = "bottomright",
    labFormat = labelFormat(suffix = " µg/m³")
  ) 

```

### 6. Choose a visualization to examine whether PM2.5 mass is associated with FEV.

```{r}
ggplot(dat, aes(x = pm25_mass, y = fev)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") + 
  labs(
    x = "Mean PM2.5 Mass (µg/m³)",
    y = "FEV",
    title = "Scatter Plot of PM2.5 Mass vs. FEV"
  ) +
  theme_minimal()

#jitter added for visualization
ggplot(dat, aes(x = pm25_mass, y = fev)) +
  geom_jitter(width = 2, height = 0.1) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +  # Add regression line
  labs(
    x = "PM2.5 Mass (µg/m³)",
    y = "FEV",
    title = "Scatter Plot of Mean PM2.5 Mass vs. FEV"
  ) +
  theme_minimal()








```

There does not seem to be a significant association between mean PM2.5 mass and FEV. The regression line is essentially horizontal meaning there is no correlation between pm2.5 mass and fev.

```{r}
cor(dat$pm25_mass, dat$fev)
cor(dat$smoke, dat$fev)
cor(dat$pm25_mass, dat$fev)
cor(dat$weight, dat$fev)


```

Here we see that weight and FEV have a correlation coefficient of 0.53. This means that as weight tends to increase, FEV also tends to increase, and vice versa, although the relationship is not extremely strong.

```{r}

oneway.test(fev ~pm25_mass,
            data=dat)
oneway.test(fev ~smoke,
            data=dat)
oneway.test(fev ~gasstove,
            data=dat)

```

In this analysis, there is no strong evidence to suggest a significant relationship between FEV and pm25_mass, smoke, or gas exposure.

## Conclusion

In this analysis of the CHS (Children's Health Study) data, we conducted a comprehensive examination of various factors related to lung function and air quality. Here are the key findings and conclusions drawn from the analysis:

Imputation was performed to handle missing values within specific groups. Continuous variables were imputed using the mean value of their respective groups, while categorical variables were imputed with the mode of their groups. Obesity level categories were created based on BMI values. Smoke and gas exposure variables were combined to create categories based on exposure status.

There is a significant positive correlation between Body Mass Index and Forced Expiratory Volume. This means that as BMI increases, FEV tends to increase as well. Weight and FEV also exhibit a significant positive correlation, indicating that as weight increases, FEV tends to increase. The analysis did not find a significant association between PM2.5 mass and FEV. The analysis did not find a significant association between smoke or gas exposure and FEV.
